Supporting Information

### 2018-02-13 manuscript.Rmd
workflow tidied; many edits especially introduction; suppinfo guidelines followed

# Notes

- bel looks for data on low PAR; maybe replace gslowA with gslowPAR, so we can argue that g0 is much higher than gslowPAR. 
- some data behind figure 4a (and maybe Arnaud 4d?)
- min(g0, gs) ('Uddling', has he used it?)  vs. g0 + gs.

We could say that plant dilemma is to be both impermeable to water loss (because they are saturated with water and leave are in an extremely dry air environment and hence need a external layer to prevent dessication= cuticle) AND need to be permeable to CO2 to grow. The dilemma is that leaves cannot be permeable to CO2 and not to H20. Hence the evolution of stomata that disrupt the cuticle layer enabling CO2 entre but cannot avoid some CO2 exit but can also close to limit H20. loss.

  It could be interested to discuss somewhere in the text the relevance of the conductance (gmin) versus the flow (Emin). Obviously gmin is more an intrinsic property of the leaf but what matters for the leaf water relation of the flow. In terms of drought resistance and water balance (and water relations) Ecuti is more relevant than gmin I think. At night for instance gdark is high but vpd so the flow might not be so different than at gmin during the day.

finding of Schuster e al (2016) who found a high proportion of triterpenoid in the cuticule of a desertic thermotolerant species. This compound would act as to stabilize the cuticule.

@xu1995 lower cuticular transpiration in droughted tomato plants, and increased wax content. Similar response to high salinity, which also reduces water availability.

@sack2003 lower gmin in sun leaves (for 2 out of 4 species), again in the same direction as evaporative demand for sun leaves is frequently higher.

@mengel1989 increased gmin in P. abies after treatment with acidic fog, the epicuticular wax layer showing a 'melted appearance'.

@@

\
Stomata exert a strong control on the transpiration rate of leaves [@bange1953]. It is well known that stomata respond quickly to changes in light intensity, humidity and CO~2~ concentration at the leaf surface. In doing so, stomata can be thought to balance the uptake of CO~2~ with the loss of water, and thus can be expected to close stomata when photosynthetic rates are low. To avoid desiccation and ultimate death, stomata can also be expected to close during periods of severe water stress. Plants cannot, however, completely halt water loss even when stomata are closed. There is increasing recognition that the minimum conductance plays an important role in estimating water fluxes in plant canopies [@barnard2013], during heat waves [@kala2016], and in models of plant drought response [@blackman2016; @martinstpaul2017].

\
Most current-generation process-based global vegetation models (GVMs) use a version of the Ball-Berry class of stomatal conductance (g~s~) models characterized by the dependence on assimilation rate, CO~2~ and humidity (Eq. 1).

@@

\
In this review, we demonstrate that inclusion of a minimum conductance is necessary both for simulation of the response of water-use efficiency to various environmental drivers, as well as a component of the response of plants to severe drought. However, there is no guidance at the moment on how to estimate minimum conductance for use in models, how various definitions and measurements compare quantitatively, whether it is feasible to assign species-specific values of the minimum conductance, and what are the ecological and environmental controls on g~min~ [@schuster2017].

@@

\
The minimum conductance of intact leaves is typically measured by mass loss of detached leaves (MLD). In this method, leaves are detached and leaf mass is monitored over time as the leaf dries out. Early work by @hygen1951 showed that after a leaf is detached, initial water loss rates are high, but after some time a constant low rate is achieved. From this minimum transpiration rate (sometimes described as the 'residual transpiration', or 'epidermal transpiration'), the minimum conductance (g~min~) can be estimated using measured vapour pressure deficit [@sinclair1986]. This method most closely resembles conditions that plants may experience during a dry-down or periods of extreme stress. We do not discuss the methodology of this measurement in detail, but note that care needs to be taken to yield reliable estimates (see SuppInfo for more discussion and illustration).

\
We quantitatively compared each of the different 'minimum' conductance (Table 1) values by synthesizing existing data. To estimate g~0~ (Eq. 1) from regression, we used the @lin2015 database to fit the model of @medlyn2011 using non-linear regression, and data presented by @miner2017 on the linear regression of the @ballberry1987 model. Measurements of conductance of isolated, non-stomatal cuticles (g~cuti~) were taken from the compilations by @kerstiens1996 and @schuster2017 (both also include g~min~ data but were not used since they included no metadata). Data on minimum conductance (g~min~) from MLD was compiled from 40 original papers (see SuppInfo 2 for references). For the compilation, we returned to all original papers mentioned in the review by @kerstiens1996, and added many newer sources (see Supporting Information for full details of the compilation). All conductance values were converted to per unit projected surface area. Although it is quite clear that night-time conductance (g~dark~) is much larger than a true 'minimum' conductance (discussed further in Section VI), it is frequently used as an independent measurement of g~0~ [@lombardozzi2017], and thus a comparison of magnitudes is useful. A compilation of night-time conductance (g~dark~) values was taken from @lombardozzi2017, who compiled cuvette (and sometimes sap flow-based) measurements of ca. 150 species during night-time. Actual minimum daytime values were estimated from the @lin2015 database, by taking those data where A~n~ < 2 $mu$ mol m^-2^ s^-1^, and calculating the minimum g~s~ for each of the 52 studies included in that dataset. This estimate was added as a reference as it represents an actual minimum values measured, rather than a fitted or extrapolated g~0~ term.

@@

\
Although we use the term cuticular conductance freely, this transport pathway does not represent a true conductance since water does not diffuse as a gas through the cuticle. Instead, it dissolves into the medium of the cuticle, diffuses through the solid matrix, and is desorped at the outer edge of the cuticle [@kerstiens1996; @riederer2001; @schreiber1996]. The main barrier to diffusion is actually a very thin layer at the leaf surface. Because most of the resistance is located in such a thin layer, g~cuti~ does not correlate with the thickness of the cuticle [@priestley1943; @riederer2001; @anfodillo2002; @schuster2016b]. Nonetheless there is considerable variation in cuticle thickness between plant species [@schuster2016a], as well as within species in response to drought (**cite**), along altitudinal transects [@delucia1984], and even with increasing height in the canopy of very tall trees [@woodruff2010]. If the cuticle thickness does not directly affect its conductance, what is the value of increasing the cuticle?

\
It is important to bear in mind the many other functions of the cuticle, including attenuation of radiation, as a barrier for fungal pathogens, various interactions with insects including signalling and herbivory resistance [@mueller2008; @kerstiens1996c;@riederer2008], and foliar water uptake [@fernandez2017] The thickness of the cuticle may also confer mechanical strength, as @onoda2012 reported that thicker cuticles are more resistant to tearing (the force to tear and tensile strength both approx proportional to cuticle thickness). As such, thick cuticles may be advantageous because prolonged exposure to wind increases the conductance of the cuticle [@hadley1983;@grace1974;@vangardingen1991], by dislodging cuticular compounds by abrasion [@rogge1993]. Similarly, exposure to simulated rain damages the cuticle and increases its conductance [@baker1986], most dramatically demonstrated in an ice storm, leading to much higher g~min~ [@boyce2003]. Such exposure to the elements was also held responsible for thicker cuticles in coastal, wind-exposed sites compared to inland sites [@blackman2005].

@@

\
In the following sections, we review the quite substantial literature on g~min~ measured with the MLD technique, stretching back to the 1930s [@pisekberger1938], focussing on environmental and ecological determinants. We also further analyze the literature compilation of g~min~ (Fig. 6). Comparing all available data, we did not find significant relationships between g~min~ and climate of origin, nor were there meaningful relationships with other traits (see Supporting Information). We thus found it difficult to explain the variation in g~min~ between species. A simple breakdown by taxonomic order (Fig. 6b) revealed that grasses (*Poales*) have a higher g~min~ compared to other orders, and conifers (especially *Pinales*) tended towards the lower range of values (but were only significantly different from *Poales*).

\
Very few studies have found meaningful correlations between g~min~ and environmental factors or ecological categories. Based on their own compilation of the literature, @schuster2017 concluded no significant differences in g~min~ or g~cuti~ (pooled) by plant growth form. One notable exception is @brodribb2014, who showed a correlation between rainfall at species origin (of the driest quarter), and g~min~. Such correlation did not bear out across all species in our database, and interesting to note is that *Eucalyptus* species (n = 11, included in *Myrtales*) - all measured on the driest continent of Australia - have g~min~ values slightly higher than the average (Fig. 6b).

\
In crop science, g~min~ has long been identified as a key drought tolerance trait [@sinclair1986]. A number of studies have targeted g~min~ as a key trait for breeding more drought-tolerant crops, leading to comparisons of g~min~ across genotypes grown in the same conditions. For example, @james2008 compared g~min~ in 58 soybean (*Glycine max*) genotypes, and found > 2-fold variation that could not be easily explained by other traits. In Fig. 6, we have compiled a number of studies in crops, demonstrating not just variation in g~min~ between crops, but also the wide range in g~min~ between genotypes. Again, it is striking that the wide variation in g~min~ cannot be easily explained by variation in leaf or other traits or chemical and structural components of the cuticle [@bengtson1978; @james2008; @saito2010].

@@

\
While there is clearly plenty of variation in g~min~ between species, a number of lines of evidence suggest that g~min~ also has great potential for plasticity. Here we summarize the literature on the acclimation of g~min~ to drought conditions, to changes in temperature and humidity, and the change in g~min~ with leaf age and with altitude. This discussion is directly relevant to the use of g~min~ in models, because if the degree of plasticity is large, it complicates model parameterization.

\
There is a general tendency for a decreased g~min~ in plants acclimated to drought stress [@james2008]. The magnitude of the decrease of g~min~ with acclimation to drought stress varied from -4 to -70% (across 10 studies, see Table S2), with a typical decrease in the order of 30-40%. In each of the studies summarized, plants were grown in well-watered or droughted conditions, and in one case a difference in g~min~ was demonstrated after just four days of drought exposure [@bengtson1978]. In a similar vein, @fanourakis2013 reported much lower g~min~ in plants grown in low vs. high relative humidity, again demonstrating that g~min~ is reduced in response to conditions that cause water limitation. We also demonstrate drought acclimation via a case study on 11 *Hakea* species grown in two watering treatments (Fig. 7) - all 11 species showed a decrease in g~min~ in the drought treatment (see SuppInfo 3 for methods details). This significant change in g~min~ with drought acclimation is likely an important component of the overall drought hardening of plants.

@@

\
The response of g~min~ to temperature is more complex; it shows both a response to instantaneous changes in temperature, as well as an acclimatory response to growth temperature. @riederer2001 and @schuster2016b have both demonstrated a steep non-linear instantaneous response of g~cuti~ to temperature, with the response becoming especially steep at higher temperatures (> 40C). In *Eucalyptus haemastoma*, the response of g~min~ to temperature was so steep, that the proportion of cuticular to total transpiration increased from 2-3% at 20C, to 40% at 38C [@eamus2008]. The mechanism of the instantaneous temperature response is complex, highly species-specific, and we refer to @schuster2016b for a detailed investigation. The quick increase in g~min~ at high temperature may well be a crucial component of the ability of plants to tolerate heat waves [@drake2018]. In support of the link between g~min~ and heat tolerance, @schuster2016a reported on a negative between thermal tolerance and g~min~ across nine species, so that species with improved tolerance to very high temperature had a low g~min~.

\
Less well established is the acclimatory potential of g~min~ to changes in growth temperature. We present a case study on *Eucalyptus parramattensis* grown in whole-tree chambers (see SuppInfo 3 for methods details). The chambers either tracked ambient conditions, or were subjected to a persistent +3C warming treatment. After several months of growth in the treatments, g~min~ was measured at various temperatures, ranging from 17.5 to 27.5C. We found a 56% decrease in g~min~ in the elevated temperature trees, but there was no clear pattern with measurement temperature (Fig. 8). This decrease in g~min~ is consistent with the drought response, because again g~min~ is reduced in leaves that are subjected to conditions that increase evaporative demand (vapour pressure deficit was higher in the elevated temperature treatment, see @drake2018). The direction of this response is consistent with @duarte2016, who reported lower g~0~ and g~dark~ in a heat wave treatment (Douglas-fir), which persisted for some time. Responses across temperature gradients are more complex, in particular the change in g~min~ with altitude has been well studied. It is commonly  reported that g~min~ increases with altitude [e.g. @delucia1984; @herrick1991; @anfodillo2002]. @fernandez2017 discuss the literature on altitude responses in detail and argue that the short growing seasons at high altitude are insufficient for complete maturation of leaf cuticles.

\
Finally we mention the striking effect of leaf age on g~min~. We have summarized five studies in Fig. S3, all of which reported on increased g~min~ for older leaves for the majority of species studied. The data reported by @jordan2007 is particularly impressive, as g~min~ in the herb *Agastachys odorata* gradually increased with leaf age up to ca. 10 years. The most likely explanation for this change in g~min~ is the continued exposure to wind, rain, and abrasives, which have been shown to damage the cuticle and increase its conductance (see Section III). In plants that maintain several cohorts of leaves, finding an appropriate value of g~min~ for use in models will have to take into account the leaf age effect.

@@

\
**Figure 8**. Minimum conductance (g~min~) measured with mass loss of detached leaves on *Eucalyptus parramattensis* grown in whole-tree chambers. Trees were grown following ambient conditions, or in a constant elevated temperature treatment (+3 degrees C) [see @drake2018]. Measurements of g~min~ were carried out at five different temperatures on replicate leaves. There was no consistent effect of measurement T on g~min~, but leaves of trees grown in elevated temperature showed on average a 45% reduction in g~min~ (p < 0.01, mixed-effects model).

@@

where g~min~ is parameterized with data obtained from mass loss of detached leaves (Fig. 6), representing a true whole-leaf minimum that is reached under drought conditions, and g~0~ is the minimum reached during low photosynthesis (due to low light or other factors) for well-watered plants. During drought, g~0~ will decrease [@misson2004], so that the true minimum reached for Eq. 1 is g~min~.

@@

>>>>

@@

\
Plants face a dilemma in constructing leaves that minimize water loss while allowing the uptake of CO~2~. Since a membrane that is permeable to CO~2~ but not H~2~O has never evolved, all land plants have stomata in their leaves, which disrupt the cuticle and allow CO~2~ uptake. It is well known that stomatal open and close in response to changes in light intensity, humidity and CO~2~ concentration at the leaf surface. In doing so, stomata can be thought to balance the uptake of CO~2~ with the loss of water, and thus can be expected to close stomata when photosynthetic rates are low. To avoid desiccation and ultimate death, stomata can also be expected to close during periods of severe water stress. When stomata are closed, water loss continues at a greatly diminished rate through the cuticle. After accounting for evaporative demand, this rate of water loss is expressed as the minimum conductance of a leaf. There is increasing recognition that the minimum conductance plays an important role in estimating water fluxes in plant canopies [@barnard2013], during heat waves [@kala2016], and in models of plant drought response [@blackman2016; @martinstpaul2017].

\
In this review, we discuss the role of the minimum conductance in models of plant water use, and critically evaluate sources of variation in this overlooked parameter. In this context, it is useful to briefly summarize the approach taken by the majority of vegetation model. Most current-generation process-based global vegetation models (GVMs) use a version of the Ball-Berry class of 'stomatal' (technically including both stomata and the cuticle) conductance (g~s~) models characterized by the dependence on assimilation rate, CO~2~ and humidity (Eq. 1).

@@

@@

\
The minimum conductance of intact leaves is typically measured by mass loss of detached leaves (MLD). In this method, leaves are detached and leaf mass is monitored over time as the leaf dries out. Early work by @hygen1951 showed that after a leaf is detached, initial water loss rates are high, but after some time a constant low rate is achieved. From this minimum transpiration rate (sometimes described as the 'residual transpiration', or 'epidermal transpiration'), the minimum conductance (g~min~) can be estimated using measured vapour pressure deficit [@sinclair1986]. This method most closely resembles conditions that plants may experience during a dry-down or periods of extreme stress. We do not discuss the methodology of this measurement in detail, but note that care needs to be taken to yield reliable estimates (see Methods S2 for further information).

\
We quantitatively compared each of the different 'minimum' conductance (Table 1) values by synthesizing existing data. To estimate g~0~ (Eq. 1) from regression, we used the @lin2015 database to fit the model of @medlyn2011 using non-linear regression, and data presented by @miner2017 on the linear regression of the @ballberry1987 model. Measurements of conductance of isolated, non-stomatal cuticles (g~cuti~) were taken from the compilations by @kerstiens1996 and @schuster2017 (both also include g~min~ data but were not used since they included no metadata). Data on minimum conductance (g~min~) from MLD was compiled from 40 original papers (see Methods S1 for description and references). For the compilation, we returned to all original papers mentioned in the review by @kerstiens1996, and added many newer sources (see Supporting Information for full details of the compilation). All conductance values were converted to per unit projected surface area. Although it is quite clear that night-time conductance (g~dark~) is much larger than a true 'minimum' conductance (discussed further in Section VI), it is frequently used as an independent measurement of g~0~ [@lombardozzi2017], and thus a comparison of magnitudes is useful. A compilation of night-time conductance (g~dark~) values was taken from @lombardozzi2017, who compiled cuvette (and sometimes sap flow-based) measurements of ca. 150 species during night-time. Actual minimum daytime values were estimated from the @lin2015 database, by taking those data where A~n~ < 2 $mu$ mol m^-2^ s^-1^, and calculating the minimum g~s~ for each of the 52 studies included in that dataset. This estimate was added as a reference as it represents an actual minimum values measured, rather than a fitted or extrapolated g~0~ term.

@@

\
Although we use the term cuticular conductance freely, this transport pathway does not represent a true conductance since water does not diffuse as a gas through the cuticle. Instead, it dissolves into the medium of the cuticle, diffuses through the solid matrix, and is desorped at the outer edge of the cuticle [@kerstiens1996; @riederer2001; @schreiber1996]. The main barrier to diffusion is actually a very thin layer at the leaf surface. Because most of the resistance is located in such a thin layer, g~cuti~ does not correlate with the thickness of the cuticle [@priestley1943; @riederer2001; @anfodillo2002; @schuster2016b]. Nonetheless there is considerable variation in cuticle thickness between plant species [@schuster2016a], along altitudinal transects [@delucia1984], and even with increasing height in the canopy of very tall trees [@woodruff2010]. If the cuticle thickness does not directly affect its conductance, what is the value of increasing the cuticle?

\
It is important to bear in mind the many other functions of the cuticle, including attenuation of radiation, as a barrier for fungal pathogens, various interactions with insects including signalling and herbivory resistance [@mueller2008; @kerstiens1996c;@riederer2008], and foliar water uptake [@fernandez2017]. The thickness of the cuticle may also confer mechanical strength, as @onoda2012 reported that thicker cuticles are more resistant to tearing (the force to tear was proportional to the cuticle thickness). As such, thick cuticles may be advantageous in exposed environments to avoid excessive damage [@blackman2005]. Prolonged exposure to wind has been shown to increase the conductance of the cuticle [@hadley1983;@grace1974;@vangardingen1991], by dislodging cuticular compounds by abrasion [@rogge1993]. Similarly, exposure to simulated rain damages the cuticle and increases its conductance [@baker1986], most dramatically demonstrated in an ice storm, leading to much higher g~min~ [@boyce2003].

@@

\
In the following sections, we review the quite substantial literature on g~min~ measured with the MLD technique, stretching back to the 1930s [@pisekberger1938], focussing on environmental and ecological determinants. We also further analyze the literature compilation of g~min~ (Fig. 6). Comparing all available data, we did not find significant relationships between g~min~ and climate of origin, nor were there meaningful relationships with other traits (see Methods S1). We thus found it difficult to explain the variation in g~min~ between species. A simple breakdown by taxonomic order (Fig. 6b) revealed that grasses (*Poales*) have a higher g~min~ compared to other orders, and conifers (especially *Pinales*) tended towards the lower range of values (but were only significantly different from *Poales*).

\
Very few studies have found meaningful correlations between g~min~ and environmental factors or ecological categories. Based on their own compilation of the literature, @schuster2017 concluded no significant differences in g~min~ or g~cuti~ (analyzed together) by plant growth form. One notable exception is @brodribb2014, who showed a correlation between rainfall at species origin (of the driest quarter), and g~min~. Such correlation did not bear out across all species in our database, and interesting to note is that *Eucalyptus* species (n = 11, included in *Myrtales*) - all measured on the driest continent of Australia - have g~min~ values slightly higher than the average (Fig. 6b).

\
In crop science, g~min~ has long been identified as a key drought tolerance trait [@sinclair1986]. A number of studies have targeted g~min~ as a key trait for breeding more drought-tolerant crops, leading to comparisons of g~min~ across genotypes grown in the same conditions. For example, @james2008 compared g~min~ in 58 soybean (*Glycine max*) genotypes, and found > 2-fold variation that could not be easily explained by other traits. In Fig. 6c, we have compiled a number of studies in crops, demonstrating not just variation in g~min~ between crops, but also the wide range in g~min~ between genotypes. Again, it is striking that the wide variation in g~min~ cannot be easily explained by variation in leaf or other traits or chemical and structural components of the cuticle [@bengtson1978; @james2008; @saito2010].

@@

\
While there is clearly plenty of variation in g~min~ between species, a number of lines of evidence suggest that g~min~ also has great potential for plasticity. Here we summarize the literature on the acclimation of g~min~ to drought conditions, to changes in temperature and humidity, and the change in g~min~ with leaf age, altitude and other factors. This discussion is directly relevant to the use of g~min~ in models, because if the degree of plasticity is large, it complicates model parameterization.

\
There is a general tendency for a decreased g~min~ in plants acclimated to drought stress [@james2008]. The magnitude of the decrease of g~min~ with acclimation to drought stress varied from -4 to -70% (across 10 studies, see Table S2), with a typical decrease in the order of 30-40%. In each of the studies summarized, plants were grown in well-watered or droughted conditions, and in one case a difference in g~min~ was demonstrated after just four days of drought exposure [@bengtson1978]. We also demonstrate drought acclimation via a case study on 11 *Hakea* species grown in two watering treatments (Fig. 7) - all 11 species showed a decrease in g~min~ in the drought treatment (see Methods S3 for experimental details). This significant change in g~min~ with drought acclimation is likely an important component of the overall drought hardening of plants.

\
The idea that water limitation causes a reduction in g~min~ (via changes in the chemical composition of the cuticle) can be tested by inspecting the response to factors that increase evaporative demand. As one of the very few studies testing this idea, @fanourakis2013 reported much lower g~min~ in *Rosa* sp. plants grown in 60% vs. 95% relative humidity. The difference could be attributed largely to a change in stomatal anatomy and lack of closure during desiccation, not changes in the cuticle *per se*. Also relevant is @sack2003, who reported lower g~min~ in sun leaves (for 2 out of 4 species) compared to shade leaves, again a lower g~min~ for leaves acclimated to high evaporative conditions.

@@

\
The response of g~min~ to temperature is more complex; it shows both a response to instantaneous changes in temperature, as well as an acclimatory response to growth temperature. @riederer2001 and @schuster2016b have both demonstrated a steep non-linear instantaneous response of g~cuti~ to temperature, with the response becoming especially steep at higher temperatures (> 40C). In *Eucalyptus haemastoma*, the response of g~min~ to temperature was so steep, that the proportion of cuticular to total transpiration increased from 2-3% at 20°C, to 40% at 38°C [@eamus2008]. The mechanism of the instantaneous temperature response is complex, highly species-specific, and we refer to @schuster2016b for a detailed investigation. The quick increase in g~min~ at high temperature may well be a crucial component of the ability of plants to tolerate heat waves [@drake2018]. In support of the link between g~min~ and heat tolerance, @schuster2016a reported on a negative between thermal tolerance and g~min~ across nine species, so that species with improved tolerance to very high temperature had a low g~min~.

\
Less well established is the acclimatory potential of g~min~ to changes in growth temperature. We present a case study on *Eucalyptus parramattensis* grown in whole-tree chambers (see Methods S3 for experimental details). The chambers either tracked ambient conditions, or were subjected to a persistent +3°C warming treatment. After several months of growth in the treatments, g~min~ was measured at various temperatures, ranging from 17.5 to 27.5°C. We found a 56% decrease in g~min~ in the elevated temperature trees, but there was no clear pattern with measurement temperature (Fig. 8). This decrease in g~min~ is consistent with the drought response, because again g~min~ is reduced in leaves that are subjected to conditions that increase evaporative demand (vapour pressure deficit was higher in the elevated temperature treatment, see @drake2018). The direction of this response is consistent with @duarte2016, who reported lower g~0~ and g~dark~ in a heat wave treatment in *Pseudotsuga menziesii*, which persisted for some time. Responses across temperature gradients are more complex, in particular the change in g~min~ with altitude has been well studied. It is commonly  reported that g~min~ increases with altitude [e.g. @delucia1984; @herrick1991; @anfodillo2002]. @fernandez2017 discuss the literature on altitude responses in detail and argue that the short growing seasons at high altitude are insufficient for complete maturation of leaf cuticles.

\
Finally we mention the striking effect of leaf age on g~min~. We have summarized five studies in Fig. S3, all of which reported on increased g~min~ for older leaves for the majority of species studied. The data reported by @jordan2007 is particularly impressive, as g~min~ in the woody shrub *Agastachys odorata* gradually increased with leaf age up to ca. 10 years. The most likely explanation for this change in g~min~ is the continued exposure to wind, rain, and abrasives, which have been shown to damage the cuticle and increase its conductance (see Section III). In plants that maintain several cohorts of leaves, finding an appropriate value of g~min~ for use in models will have to take into account the leaf age effect.

@@

\
**Figure 8**. Minimum conductance (g~min~) measured with mass loss of detached leaves on *Eucalyptus parramattensis* grown in whole-tree chambers. Trees were grown following ambient conditions, or in a constant elevated temperature treatment (+3 °C) [see @drake2018]. Measurements of g~min~ were carried out at five different temperatures on replicate leaves. There was no consistent effect of measurement temperature on g~min~, but leaves of trees grown in elevated temperature showed on average a 45% reduction in g~min~ (p < 0.01, mixed-effects model).

@@

where g~min~ is parameterized with data obtained from mass loss of detached leaves (Fig. 6), representing a true whole-leaf minimum that is reached under conditions of severe drought, and g~0~ is the minimum reached during low photosynthesis (due to low light or other factors) for well-watered plants. During drought, g~0~ will decrease [@misson2004], so that the true minimum reached for Eq. 1 is g~min~.

@@

# Supporting Information

**Methods S1** Literature compilation of g~min~ estimates
**Methods S2** Methodological issues with measurement of g~min~
**Methods S3** Methods description for case studies
**Table S1** Effects of drought stress on g~min~
**Figure S1** Effects of leaf age on g~min~

### 2018-02-13 manuscript_suppinfo.Rmd
workflow tidied; many edits especially introduction; suppinfo guidelines followed

# 1. Literature compilation of g~min~ estimates

@@

```{r gminfig1, echo=FALSE, dpi=300, fig.width=6, fig.height=5}
figure_gmin_bygroup(gmindat)
```
**Figure S1**. Average minimum conductance (g~min~) for crops, graminoids, woody and other non-woody plant species. Grey symbols are original datapoints. Numbers above the figure refer to the number of species in each group. Different letters indicate significant differences. Error bars are 95% confidence intervals for the mean (calculated for log-transformed g~min~, and back-transformed).

```{r gminfig2, echo=FALSE, dpi=300, fig.width=6, fig.height=5}
figure_pet(gmindat2)
```
**Figure S2**. No relationship between g~min~ and potential evapotranspiration across the range of the species. Climate variables were calculated across the distribution of each species; we tested many other variables (precipitation, temperature, etc.), but no relationship with climate could be determined.

# 2. Methodological issues with measurement of g~min~

@@

```{r goodbadfig, echo=FALSE, dpi=300, fig.width=8, fig.height=4}
figure_goodandbad(goodbadcurves)
```
**Figure S3**. Two example curves for determination of minimum conductance (g~min~) via mass loss of detached leaves (MLD). Leaf weights are expressed as relative water content (RWC). The good quality curve shows a distinct linear portion at the end of the drying curve, from which minimum transpiration can be calculated (using actual leaf weight, and leaf area). The poor quality curve shows a quicker decrease towards the end of the drying curve (> 400min), indicating an *increased* transpiration rate despite the low water content. We speculate that this increase - which we have observed in a number of curves - is due to the epidermis drying so quickly that the stomata are 'pulled open', increasing water loss.

# 3. Methods description for case studies

## Drought acclimation in *Hakea*

@@

\
We determined g~min~ from detached leaves collected before sunrise from each of 6 *Eucalyptus parramattensis* trees grown under ambient and ambient+3C temperature conditions, respectively. Leaves were scanned for leaf area determination before being weighed using a precision balance and attached to a ‘clothesline’ suspended inside controlled chambers set at different temperatures (17.5, 20, 22.5, 25, 27.5°C) and common low light (~200 umol m^-2^ s^-1^) and a relative humidity of 60%. Inbuilt fans in each chamber blew air gently across the leaf surfaces. Leaves were allowed to slowly dehydrate over a period of 7-8 hours during which time leaf mass was measured roughly every hour. For each leaf, transpiration was calculated from the slope of the linear part of the leaf mass vs time relationship, and converted to g~min~.

# 4. Effects of drought stress on g~min~

@@

# 5. Effects of leaf age on g~min~

@@

**Figure S4**. Correlation between g~min~ and leaf age for five studies in the g~min~ database. Leaf age is expressed in years (0 = current year), with the exception of @phillips2010, who measured juvenile and adult foliage on *Eucalyptus*. Citations with species studied are @boyce2000 (*Abies lasiocarpa*, *Picea engelmanii*), @boyce2003 (*Tsuga canadensis*, *Pinus strobus*, *Pinus resinosa*, *Picea rubens*), @heinsoo1998 (*Picea abies*), @jordan2007 (*Agastachys odorata*), @phillips2010 (*Eucalyptus saligna*, *E. sideroxylon*, *E. cladocalyx*).

>>>>

# Methods S1. Literature compilation of g~min~ estimates

@@

```{r gminfig1, echo=FALSE, dpi=300, fig.width=6, fig.height=5}
figure_gmin_bygroup(gmindat)
```
**Figure 1**. Average minimum conductance (g~min~) for crops, graminoids, woody and other non-woody plant species. Grey symbols are original datapoints. Numbers above the figure refer to the number of species in each group. Different letters indicate significant differences. Error bars are 95% confidence intervals for the mean (calculated for log-transformed g~min~, and back-transformed).

```{r gminfig2, echo=FALSE, dpi=300, fig.width=6, fig.height=5}
figure_pet(gmindat2)
```
**Figure 2**. No relationship between g~min~ and potential evapotranspiration across the range of the species. Climate variables were calculated across the distribution of each species; we tested many other variables (precipitation, temperature, etc.), but no relationship with climate could be determined.

# Methods S2. Methodological issues with measurement of g~min~

@@

```{r goodbadfig, echo=FALSE, dpi=300, fig.width=8, fig.height=4}
figure_goodandbad(goodbadcurves)
```

**Figure 3**. Two example curves for determination of minimum conductance (g~min~) via mass loss of detached leaves (MLD). Leaf weights are expressed as relative water content (RWC). The good quality curve shows a distinct linear portion at the end of the drying curve, from which minimum transpiration can be calculated (using actual leaf weight, and leaf area). The poor quality curve shows a quicker decrease towards the end of the drying curve (> 400min), indicating an *increased* transpiration rate despite the low water content. We speculate that this increase - which we have observed in a number of curves - is due to the epidermis drying so quickly that the stomata are 'pulled open', increasing water loss.

# Methods S3. Methods description for case studies

## Drought acclimation in *Hakea* spp.

@@

\
We determined g~min~ from detached leaves collected before sunrise from each of 6 *Eucalyptus parramattensis* E.C. Hall trees grown under ambient and ambient+3C temperature conditions, respectively. Leaves were scanned for leaf area determination before being weighed using a precision balance and attached to a ‘clothesline’ suspended inside controlled chambers set at different temperatures (17.5, 20, 22.5, 25, 27.5°C) and common low light (~200 umol m^-2^ s^-1^) and a relative humidity of 60%. Inbuilt fans in each chamber blew air gently across the leaf surfaces. Leaves were allowed to slowly dehydrate over a period of 7-8 hours during which time leaf mass was measured roughly every hour. For each leaf, transpiration was calculated from the slope of the linear part of the leaf mass vs time relationship, and converted to g~min~.

# Table S1. Effects of drought stress on g~min~

@@

# Figure S1. Effects of leaf age on g~min~

@@

**Figure S1**. Correlation between g~min~ and leaf age for five studies in the g~min~ database. Leaf age is expressed in years (0 = current year), with the exception of @phillips2010, who measured juvenile and adult foliage on *Eucalyptus*. Citations with species studied are @boyce2000 (*Abies lasiocarpa*, *Picea engelmanii*), @boyce2003 (*Tsuga canadensis*, *Pinus strobus*, *Pinus resinosa*, *Picea rubens*), @heinsoo1998 (*Picea abies*), @jordan2007 (*Agastachys odorata*), @phillips2010 (*Eucalyptus saligna*, *E. sideroxylon*, *E. cladocalyx*).
